---
title: "What happened to the streak? rle: run length encoding"
author: "Richard Careaga"
date: "2019-12-26"
slug: run length encoding
output: html_document
categories:
  - Data Science
tags:
  - R
  - run length encoding
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A common task is to isolate records when some condition changes. Failure analysis is an obvious example. When working in a `tidy` environment, this isn't always easy because of its strong column bias. Looking back at previous rows is clunky, at best.

The sometimes overlooked `which()` provides one part of the solution. Consider a simple case of two vectors, easily combinable into a data frame, one of which indicates a classification and the other, whether or not that classification is present.

```{r, echo=TRUE}
genres <- c("Action", "Animation", "Comedy", "Documentary", "Romance", "Short")
row_vector <- c(0,0,1,1,0,0,0)
indices <-min(which(row_vector == TRUE))
indices
```

Both `genres` and `row_vector` are equal-length character and numeric vectors and the correspondence depends on position, with the `Action 0` pair being the first and `Romance 0` the last. Python users will recognize this as a `hash`.

The `indices` atomic (single element) vector applies, working from the inside out, the `which()` function to select the elements of `row_vector` that are `TRUE`, which evaluates to `1` and `min` finds the position of the first among that subset. So, we end up with `3` and `genres[3]` evaluates to **`r genres[3]`**.

Using the positions of elements in one vector to identify elements in another provides a way to use `rle()`, the run-length encoding function.

What `rle` does is to keep track of the number of times an element in a vector appears repeated zero or more times.

As usual, it helps to run the `help()` example, with some inspection:

```{r}
x <- rev(rep(6:10, 1:5))
x
y <- rle(x)
y
str(y)
```

The two pieces of `y`, `y$lengths` (or `y[1]`) and `y$values` tell us that there are five repetitions of 10, four of 9, etc.

Let's create a simple data frame

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tibble)) 
my_data <- tibble(type=rep(c(1:2), each= 9), hour= rep(1:9, 2), event = c(1,1,1,0,0,0,0,1,1,1,1,0,0,0,1,0,1,0))
my_data
```

Let `1` in column `event` indicate success and `0` failure. Where  does each string of successes turn into failure?

```{r}
runs <- rle(my_data$event)
runs <- tibble(runs$lengths, runs$values)
colnames(runs) <- c("lengths", "values")
runs
```

```{r}
sequences <- sequences <- tibble(lengths = runs$lengths, values = runs$values) %>% mutate(indices = cumsum(runs$lengths))
sequences
post_zero <- sequences %>%  filter(values == 0)
post_zero
result <- left_join(sequences, post_zero, by = "indices") %>% select(1:3) %>% filter(values.x == 1)
colnames(result) <- c("lengths", "runs", "indices")
result
my_data[result$indices,]
```


The variable `type = 1` had one string of successes that ended at hour three, `type = 2` had three ending at hours two, six and eight.

More interesting, of course, is the case where `hour` is a `datetime` object and you can bring date arithmetic into play. 

The main point is that if you can design a logical test to mutate a numeric column, `rle` provides a straightforward way of subsetting sequences based on the the test result.
