---
title: "Metadata for datasets"
author: Richard Careaga
date: "2019-10-18"
slug:  Metadata for datasets
categories:
  - Data Science
tags:
  - R
  - documentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pander)
```

# Where does this dataset come from?

> Is this the original revised data or the revised revised data?

Keeping track of the provenance of data can be a challenge, especially when drawing on published sources. Keeping a record of the origin, the date accessed, the transformations applied (e.g., converting from .xls to cvs and converting character strings such as “$1,250,321.21” to floats or date strings to date objects), subsequent changes, who handled the data object and where it can be found in a repository are all things that enhance the analyst’s own ability to reproduce results.

Unfortunately, notes go missing, files get mis-filed and all the other hazards that can befall research can happen. Often, one wishes for R objects with built-in metadata for that purpose.

# Using mostattributes() to do attach metadata

Scott Chamerlain at ropensci.org brought **attr** to my attention, which is the built-in way I was looking for originally. He also pointed me to EML, a much more elaborate approach suited for publication projects.

## A minimal example

### Create data frame and a separate metadata list

```{r, results="asis", echo = FALSE, warning=FALSE}
fips <- read.csv("https://tuva.s3-us-west-2.amazonaws.com/state_fips_postal.csv", header = FALSE)
    colnames(fips) = c("state", "fip", 'id')
    require(jsonlite) # easier to use JSON to write metadata
    meta <- fromJSON("https://tuva.s3-us-west-2.amazonaws.com/2015-07-31-meta.json")
```

    fips <- read.csv("https://tuva.s3-us-west-2.amazonaws.com/state_fips_postal.csv", header = FALSE)
    colnames(fips) = c("state", "fip", 'id')
    require(jsonlite) # easier to use JSON to write metadata
    meta <- fromJSON("https://tuva.s3-us-west-2.amazonaws.com/2015-07-31-meta.json")

The json source file looks like this

    [
        {
        "Accessed": "2015-07-31",
        "GitBlame": "Richard Careaga",
        "Contact": "technocrat@twitter",
        "Preprocessing": "FIPS Codes for the States and District of Columbia table captured manually and converted to cvs file",
        "Source": "https://www.census.gov/geo/reference/ansi_statetables.html",
        "Repository": "unassigned",
        "Version": "1.0"
        }
    ]

### Associate the metadata with the data frame using mostattributes

    x <- fips
    mostattributes(x) <- list(meta = meta)

```{r, results="asis", echo = FALSE, warning=FALSE}
    x <- fips
    mostattributes(x) <- list(meta = meta)
``` 

### Now metadata is displayed by default

```{r}
x
```


